### 2021-04-13 paper/ase21/analyzer.tex
[ASE21] Till 4.2

In this section, we explain the detail of the type analyzer of $\tool$. We first
formally define a modified $\ires$ and explain how to perform type analysis for
the modified $\ires$.  Moreover, we also explain a refinement of type analysis
based on conditions of assertions and branches to increase analysis precision.

@@

$\ires$ is an untyped intermediate representation introduced by \citet{jiset}.
To make it more suitable for type analysis, we modify it as a label-based
language as follows:
\small
\[
  \begin{array}{l@{~}r@{~}c@{~}r@{~}r@{~}l}
    \text{Functions}
    &\funcset&\ni&\func&::=&\kwdef \; \x (\x^*, \x^*). \; \lab\\

@@

A modified $\ires$ program $\prog = (\getfunc, \getinst, \getnext)$ is consists
of three mappings from labels;  $\getfunc: \labset \rightarrow \funcset$
represents that labels are in which functions, $\getinst: \labset \rightarrow
\instset$ represents instructions attached to labels, and $\getnext: \labset
\rightarrow \labset$ denotes next labels where a label $\lab \in \labset$
denotes a program point.  function $\kwdef \; \f (\x^*, \y^*). \; \lab \in
\funcset$ consists of its name $\f$, normal parameters $\x^*$, optional
parameters $\y^*$, and a body label $\lab$.  Although global variables exist in
ECMAScript, we assume that only local variables exist in this paper for brevity
An instruction $\inst$ is a variable declaration, a function call, an assertion,
a branch, a return, or an reference update.  Each invocation of an abstract
algorithm in ECMAScript is compiled to a function call instruction with a newly
introduced temporal variable.  We represent loops using branch instructions with
cyclic pointing of labels in the next mapping $\getnext$.  A reference $\refer$
is a variable $\x$ or a field access $\refer \kwsl \expr \kwsr$.  An expression
$\expr$ is a record, a list, a type check, an existence check, a binary
operation, a unary operation, a reference, a constant, or a primitive, which is
either $\undefval$, $\nullval$, a Boolean $\bool$, a Number $\num$, a BigInt
$\bigint$, a String $\str$, or a Symbol $\symb$.

@@

A type $\ty \in \tyset$ is a norminal type $\tname$ or a pre-defined type;
$\tjs$ denotes JavaScript values, $\tprim$ primitives, $\tnumeric$ either Number
or BigInt values, and $\tnum$, $\tbigint$, $\tstr$, and $\tsymb$ denote Number,
BigInt, String, and Symbol values, respectively.  A norminal type $\tname$ is
either 1) a \textit{production type} for ASTs or 2) a \textit{record type} used
in abstract algorithms, and ECMAScript describes its fields with pre-defined or
possible values.  A production type has corresponding syntax-directed algorithms
as its fields and a record type has specific fields with possible values
described in the corresponding table and additional description in ECMAScript.
The production type is automatically extracted but we manually model fields of
record types.  For example, Figure~\label{fig:record-fields-table} is an excerpt
from ES11 that describes fields of completion
records~\footnote{https://262.ecma-international.org/11.0/\#table-8}; it
contains three fields ``Type'' with a constant such as \code{normal}, ``Value''
with JavaScript values or an constant \code{empty}, and ``Target'' with String
values or \code{empty}.  We model them as follows:
\small
\[
  \begin{array}{l}
    \text{Completion} = \kwcl\\
    \quad \text{Type}: \kwcl \nconst{normal}, \nconst{break}, \nconst{continue},
    \nconst{return}, \nconst{throw} \kwcr\\
    \quad \text{Value}: \kwcl \tjs, \nconst{empty} \kwcr\\
    \quad \text{Target}: \kwcl \tstr, \nconst{empty} \kwcr\\
    \kwcr\\
  \end{array}
\]

The subtype relation $\subtype \subseteq \tyset \times \tyset$ between types is
described in Figure~\ref{fig:subtype}; the directed edge from $\ty'$ to $\ty$
denotes $\ty' \subtype \ty$ and the relation is reflexive and transitive.  The
subtype relation is dependent on the syntax and defined record types in
ECMAScript.  We extract subtypes for production types and manually model
subtypes between record types.  For example, consider the syntax-directed
abstract algorithm at the lower-right in Figure~\ref{fig:example}.  The
non-terminal \textit{BindingElement} is the unique token in a alternative of
\textit{FormalParameter} production thus we automatically extract the subtype
relation: \textit{BindingElement} $\subtype$ \textit{FormalParameter}.  Based on
this subtype relation, a type check expression $\expr: \ty$ checks whether the
evaluation result of $\expr$ has a type $\ty'$ satisfying $\ty' \subtype \ty$.

We define a denotational semantics of the modified $\ires$ for instructions
$\sem{\inst}: \dom \rightarrow \dom$ and expressions $\sem{\expr}: \dom
\rightarrow \valset$.  For presentation brevity, we omit it in this paper and
include it in a companion report~\inred{\cite{report}}.
% \[
%   \begin{array}{l@{~}r@{~}c@{~}r@{~}r@{~}l}
%     \text{States}
%     &\elem&\in&\dom&::=&\ctxtset^* \times \labset \times \envset\\
% 
%     \text{Contexts}
%     &\ctxt$\in$\ctxtset&::=&\labset \times \varset \times \envset\\
% 
%     \text{Environments}
%     &\env$\in$\envset&::=&\varset \finmap \valset\\
% 
%     \text{Values}
%     &\val$\in$\valset&::=& \addrset \uplus \addr \uplus \primset\\\
%   \end{array}
% \]

@@

        \asemi{\x = \kwrl \expr_0 \; \expr_1 \cdots \expr_n \kwrr}
        (\lab, \tys)(\aelem) &=&
        \left(
          \begin{array}{@{}l@{}}
            \{ (\lab_\func, \tys_a) \mapsto \aenv_{\func, \tys_a} \mid \func \in
              F \wedge \tys_a \in T_a
            \},\\
            \{ (\func, \tys_a) \mapsto \{ (\getnext(\lab), \tys, \x) \} \mid
              \func \in F \wedge \tys_a \in T_a
            \}\\
          \end{array}
        \right) &
        \text{where} \;
        \left\{
          \begin{array}{@{}l@{}}
            \aty_i = \aseme{\expr_i}(\aenv) \wedge T_a = \{ \upcasts([ \ty_1,
            \cdots, \ty_n ] \mid \ty_i \in \aty_i \} \wedge\\
            F = \{ \func \in \aty_0 \mid \func = \kwdef \; \_ (\p_1, \cdots,
            \p_m). \; \lab_\func \} \wedge\\
            \aenv_{\func, \tys_a} = \aenv[\p_1 \mapsto \{ \tys_a[1] \}, \cdots,
            \p_m \mapsto \{ \tys_a[m] \}]\\
          \end{array}
        \right.\\

        \asemi{\kwassert \; \expr}
        (\lab, \tys)(\aelem) &=&
        (\{ (\getnext(\lab), \tys) \mapsto \aenv_\mt \}, \emp) &
        \text{where} \; \aenv_\mt = \refine(\aenv, \expr, \true)\\

        \asemi{\kwif \; \expr \; \lab_\mt \; \lab_\mf}
        (\lab, \tys)(\aelem) &=&
        (\{ (\lab_\mt, \tys) \mapsto \aenv_\mt, (\lab_\mf, \tys) \mapsto
        \aenv_\mf \}, \emp) &
        \text{where} \;
        \aenv_\mt = \refine(\aenv, \expr, \true) \wedge
        \aenv_\mf = \refine(\aenv, \expr, \false)\\

@@

    \text{Abstract Types}
    &\aty&\in&\atyset&=&\partsof{\tyset}\\
  \end{array}
\]
An abstract state $\aelem \in \adom$ is pair of a return map and a return point
mapping.  A result map $\rmap \in \rmapset$ represents an abstract environment
for each flow-sensitive and type-sensitive view, and a return point mapping
$\retp \in \retpset$ represents possible return points of each function with
a type-sensitive context.  An abstract environment $\aenv \in \aenvset$
represents possible types of each variable, and $\aenv(\x) = \{ \tabsent \}$
when $\x$ is not defined in $\aenv$.  An abstract type $\aty \in \atyset$ is a
set of types.  While the join operator $\join$ and the partial order $\order$
for abstract states are defined in a point-wise manner, the operators for types
are defined with a normalization function $\norm$ because of their subtype
relations as follows:
\[
  \begin{array}{l}
    \aty_0 \join \aty_1 = \norm(\aty_0 \cup \aty_1)\\
    \aty_0 \order \aty_1 \Leftrightarrow \forall \ty_0 \in \aty_0. \; \exists
    \ty_1 \in \norm(\aty_1). \; \text{s.t.} \; \ty_0 \subtype \ty_1\\
  \end{array}
\]
where $\norm(\aty) = \{ \ty \mid \ty \in \aty_e \wedge \not\exists \ty' \in
\aty_e. \; \text{s.t.} \; \ty \neq \ty' \wedge \ty \subtype \ty' \}$.

Based on abstract states, we define the abstract semantics of instructions,
references, and expressions in the modified $\ires$ described in
Figure~\ref{fig:abs-sem}.  To prevent the explosion of the number of type-sensitive
views, we upcast the type of each argument before function calls with the
following function:
\[
  \upcast(\ty) = \left\{
    \begin{array}{ll}
      \tnormal(\upcast(\ty')) & \text{if} \; \ty = \tnormal(\ty')\\
      \clist{\upcast(\ty')} & \text{if} \; \ty = \clist{\ty'}\\
      \tstr & \text{if} \; \ty = \str\\
      \tbool & \text{if} \; \ty = \bool\\
      \ty & \text{otherwise}\\
    \end{array}
  \right.
\]
and $\upcasts$ denotes a point-wise extension of $\upcast$ for type sequences.
For conditions, we apply the $\refine$ function that prunes out infeasible parts
in abstract states based on given conditions to increase analysis precision.  We
explain its detail in Section~\ref{sec:refine}.  Then, we define the abstract
semantics $\asem{\prog}$ of a program $\prog = (\getfunc, \getinst, \getnext)$
as the least fixpoint of abstract transfer $\atransfer: \adom \rightarrow \adom$
as follows:
\[
  \begin{array}{r@{~}c@{~}l}
    \asem{\prog} &=& \lim_{n \rightarrow \infty}(\atransfer)^n(\iaelem)\\
    \atransfer(\aelem) &=& \aelem \join \left(
      \underset{(\lab, \tys) \in \domain{\rmap}}{\bigjoin}
      \asemi{\getinst(\lab)}(\lab, \tys)(\aelem)
    \right)\\
  \end{array}
\]
where $\aelem = (\rmap, \_)$ and $\iaelem$ denotes the initial abstract state.
As described in the Section~\ref{sec:overview}, $\iaelem$ contains all
syntax-directed algorithms having no additional parameters and built-in
algorithms with appropriate abstract environments.  For syntax-directed
algorithms, we construct its abstract environment containing the variable
$\this$ with a production type and other variables for non-terminals.  For
example, the upper-left syntax-directed algorithm in Figure~\ref{fig:example} is
initialized with the following abstract environment:
\[
  \left\{
    \begin{array}{r@{~}c@{~}l}
      \this &\mapsto& \{ \text{AssignmentExpression} \},\\
      \mcode{LeftHandSideExpression} &\mapsto& \{ \text{LeftHandSideExpression} \},\\
      \mcode{AssignmentExpression} &\mapsto& \{ \text{AssignmentExpression} \}
    \end{array}
  \right.
\]
For built-in algorithms, we assign pre-defined variables $\this$, $\args$, and
$\NewTarget$ with corresponding types and parameters with $\tjs$ types.  For
example, the following abstract environment is for the built-in
algorithm $\jscode{Math.round}$ at the upper-right in Figure~\ref{fig:example}:
\[
  \left\{
    \begin{array}{r@{~}c@{~}l}
      \this &\mapsto& \{ \tjs \},
      \args \mapsto \{ \clist{\tjs} \},
      \x \mapsto \tjs\\
      \NewTarget &\mapsto& \{ \text{Object}, \undefval \},\\
    \end{array}
  \right.
\]

\subsection{Refinement}\label{sec:refine}

\todo

>>>>

In this section, we explain the detail of the type analyzer of $\tool$. We first
formally define a modified $\ires$ and explain how to perform type analysis for
the modified $\ires$.  Moreover, we also explain a condition-based refinement of
type analysis based on conditions of assertions and branches to increase
analysis precision.

@@

$\ires$ is an untyped intermediate representation introduced by \citet{jiset}.
To make it more suitable for type analysis, we modify it as a label-based
language as follows:
\small
\[
  \begin{array}{l@{~}r@{~}c@{~}r@{~}r@{~}l}
    \text{Functions}
    &\funcset&\ni&\func&::=&\kwdef \; \x (\x^*, \kwsl \x^* \kwsr). \; \lab\\

@@

A modified $\ires$ program $\prog = (\getfunc, \getinst, \getnext)$ is consists
of three mappings from labels;  $\getfunc: \labset \rightarrow \funcset$
represents that labels are in which functions, $\getinst: \labset \rightarrow
\instset$ represents instructions attached to labels, and $\getnext: \labset
\rightarrow \labset$ denotes next labels where a label $\lab \in \labset$
denotes a program point.  function $\kwdef \; \f (\x^*, \kwsl \y^* \kwsr). \;
\lab \in \funcset$ consists of its name $\f$, normal parameters $\x^*$, optional
parameters $\y^*$, and a body label $\lab$.  Although global variables exist in
ECMAScript, we assume that only local variables exist in this paper for brevity
An instruction $\inst$ is a variable declaration, a function call, an assertion,
a branch, a return, or an reference update.  Each invocation of an abstract
algorithm in ECMAScript is compiled to a function call instruction with a newly
introduced temporal variable.  We represent loops using branch instructions with
cyclic pointing of labels in the next mapping $\getnext$.  A reference $\refer$
is a variable $\x$ or a field access $\refer \kwsl \expr \kwsr$.  We use the
notation $\refer.\f$ to briefly represent $\refer \kwsl \code{"f"} \kwsr$. An
expression $\expr$ is a record, a list, a type check, an existence check, a
binary operation, a unary operation, a reference, a constant, or a primitive,
which is either $\undefval$, $\nullval$, a Boolean $\bool$, a Number $\num$, a
BigInt $\bigint$, a String $\str$, or a Symbol $\symb$.

@@

A type $\ty \in \tyset$ is a norminal type $\tname$ or a pre-defined type;
$\tjs$ denotes JavaScript values, $\tprim$ primitives, $\tnumeric$ either Number
or BigInt values, and $\tnum$, $\tbigint$, $\tstr$, and $\tsymb$ denote Number,
BigInt, String, and Symbol values, respectively.  A norminal type $\tname$ is
either 1) a \textit{production type} for ASTs or 2) a \textit{record type} used
in abstract algorithms, and ECMAScript describes its fields with pre-defined or
possible values.  A production type has corresponding syntax-directed algorithms
as its fields and a record type has specific fields with possible values
described in the corresponding table and additional description in ECMAScript.
The production type is automatically extracted but we manually model fields of
record types.  For example, Figure~\label{fig:record-fields-table} is an excerpt
from ES11 that describes fields of completion
records~\footnote{https://262.ecma-international.org/11.0/\#table-8}; it
contains three fields ``Type'' with a constant such as \code{normal}, ``Value''
with JavaScript values or an constant \code{empty}, and ``Target'' with String
values or \code{empty}.  We manually model them as follows:
\small
\[
  \begin{array}{l}
    \text{Completion} = \kwcl\\
    \quad \Type: \kwcl \nconst{normal}, \nconst{break}, \nconst{continue},
    \nconst{return}, \nconst{throw} \kwcr\\
    \quad \Value: \kwcl \tjs, \nconst{empty} \kwcr\\
    \quad \Target: \kwcl \tstr, \nconst{empty} \kwcr\\
    \kwcr\\
  \end{array}
\]

The subtype relation $\subtype \subseteq \tyset \times \tyset$ between types is
depicted in Figure~\ref{fig:subtype}; the directed edge from $\ty'$ to $\ty$
denotes $\ty' \subtype \ty$ and the relation is reflexive and transitive.  The
subtype relation is dependent on the syntax and defined record types in
ECMAScript.  We extract subtypes for production types and manually model
subtypes between record types.  For example, consider the syntax-directed
abstract algorithm at the lower-right in Figure~\ref{fig:example}.  The
non-terminal \textit{BindingElement} is the unique token in a alternative of
\textit{FormalParameter} production thus we automatically extract the subtype
relation: \textit{BindingElement} $\subtype$ \textit{FormalParameter}.  Based on
this subtype relation, a type check expression $\expr: \ty$ checks whether the
evaluation result of $\expr$ has a type $\ty'$ satisfying $\ty' \subtype \ty$.

We define a denotational semantics of the modified $\ires$ for instructions
$\semi{\inst}: \dom \rightarrow \dom$, references $\semr{\refer}: \dom
\rightarrow \valset$, and expressions $\sem{\expr}: \dom \rightarrow \valset$.
For presentation brevity, we omit it in this paper and include it in a companion
report~\inred{\cite{report}}.

@@

        \asemi{\x = \kwrl \expr_0 \; \expr_1 \cdots \expr_n \kwrr}
        (\lab, \tys)(\aelem) &=&
        \left(
          \begin{array}{@{}l@{}}
            \{ (\lab_\func, \tys_a) \mapsto \aenv_{\func, \tys_a} \mid \func \in
              F \wedge \tys_a \in T_a
            \},\\
            \{ (\func, \tys_a) \mapsto \{ (\getnext(\lab), \tys, \x) \} \mid
              \func \in F \wedge \tys_a \in T_a
            \}\\
          \end{array}
        \right) &
        \text{where} \;
        \left\{
          \begin{array}{@{}l@{}}
            \aty_i = \aseme{\expr_i}(\aenv) \wedge T_a = \{ \upcasts([ \ty_1,
            \cdots, \ty_n ] \mid \ty_i \in \aty_i \} \wedge\\
            F = \{ \func \in \aty_0 \mid \func = \kwdef \; \_ (\p_1, \cdots,
            \kwsl \cdots, \p_m \kwsr). \; \lab_\func \} \wedge\\
            \aenv_{\func, \tys_a} = \aenv[\p_1 \mapsto \{ \tys_a[1] \}, \cdots,
            \p_m \mapsto \{ \tys_a[m] \}]\\
          \end{array}
        \right.\\

        \asemi{\kwassert \; \expr}
        (\lab, \tys)(\aelem) &=&
        (\{ (\getnext(\lab), \tys) \mapsto \aenv_\mt \}, \emp) &
        \text{where} \; \aenv_\mt = \refine(\expr, \true)(\aenv)\\

        \asemi{\kwif \; \expr \; \lab_\mt \; \lab_\mf}
        (\lab, \tys)(\aelem) &=&
        (\{ (\lab_\mt, \tys) \mapsto \aenv_\mt, (\lab_\mf, \tys) \mapsto
        \aenv_\mf \}, \emp) &
        \text{where} \;
        \aenv_\mt = \refine(\expr, \true)(\aenv) \wedge
        \aenv_\mf = \refine(\expr, \false)(\aenv)\\

@@

    \text{Abstract Types}
    &\aty&\in&\atyset&=&\partsof{\tyset}\\
  \end{array}
\]
An abstract state $\aelem \in \adom$ is pair of a return map and a return point
mapping.  A result map $\rmap \in \rmapset$ represents an abstract environment
for each flow-sensitive and type-sensitive view, and a return point mapping
$\retp \in \retpset$ represents possible return points of each function with
a type-sensitive context.  An abstract environment $\aenv \in \aenvset$
represents possible types of each variable, and $\aenv(\x) = \{ \tabsent \}$
when $\x$ is not defined in $\aenv$.  An abstract type $\aty \in \atyset$ is a
set of types.  While we define the join operator $\join$, the meet operator
$\meet$, and the partial order $\order$ for most of abstract domains in a
point-wise manner, we define the operators for types with a normalization
function $\norm$ because of their subtype relations:
\[
  \begin{array}{l}
    \aty_0 \join \aty_1 = \norm(\aty_0 \cup \aty_1)\\
    \aty_0 \meet \aty_1 = \norm(
      \{\ty_0 \in \aty_0 \mid \{ \ty_0 \} \order \aty_1 \} \cup
      \{\ty_1 \in \aty_1 \mid \{ \ty_1 \} \order \aty_0 \}
    )\\
    \aty_0 \order \aty_1 \Leftrightarrow \forall \ty_0 \in \aty_0. \; \exists
    \ty_1 \in \norm(\aty_1). \; \text{s.t.} \; \ty_0 \subtype \ty_1\\
  \end{array}
\]
where $\norm(\aty) = \{ \ty \mid \ty \in \aty_e \wedge \not\exists \ty' \in
\aty_e. \; \text{s.t.} \; \ty \neq \ty' \wedge \ty \subtype \ty' \}$.

Based on abstract states, we define the abstract semantics of instructions,
references, and expressions in the modified $\ires$ described in
Figure~\ref{fig:abs-sem}.  To prevent the explosion of the number of type-sensitive
views, we upcast the type of each argument before function calls with the
following function:
\[
  \upcast(\ty) = \left\{
    \begin{array}{ll}
      \tnormal(\upcast(\ty')) & \text{if} \; \ty = \tnormal(\ty')\\
      \clist{\upcast(\ty')} & \text{if} \; \ty = \clist{\ty'}\\
      \tstr & \text{if} \; \ty = \str\\
      \tbool & \text{if} \; \ty = \bool\\
      \ty & \text{otherwise}\\
    \end{array}
  \right.
\]
and $\upcasts$ denotes a point-wise extension of $\upcast$ for type sequences.
For branches and assertions, we apply the $\refine$ function that prunes out
infeasible parts in abstract environments based on their conditions to increase
analysis precision.  We explain its detail in Section~\ref{sec:refine}.  Then,
we define the abstract semantics $\asem{\prog}$ of a program $\prog = (\getfunc,
\getinst, \getnext)$ as the least fixpoint of abstract transfer $\atransfer:
\adom \rightarrow \adom$:
\[
  \begin{array}{r@{~}c@{~}l}
    \asem{\prog} &=& \lim_{n \rightarrow \infty}(\atransfer)^n(\iaelem)\\
    \atransfer(\aelem) &=& \aelem \join \left(
      \underset{(\lab, \tys) \in \domain{\rmap}}{\bigjoin}
      \asemi{\getinst(\lab)}(\lab, \tys)(\aelem)
    \right)\\
  \end{array}
\]
where $\aelem = (\rmap, \_)$ and $\iaelem$ denotes the initial abstract state.
As described in the Section~\ref{sec:overview}, $\iaelem$ contains all
syntax-directed algorithms having no additional parameters and built-in
algorithms with appropriate abstract environments.  For syntax-directed
algorithms, we construct its abstract environment containing the variable
$\this$ with a production type and other variables for non-terminals.  For
example, the upper-left syntax-directed algorithm in Figure~\ref{fig:example} is
initialized with the following abstract environment:
\[
  \left\{
    \begin{array}{r@{~}c@{~}l}
      \this &\mapsto& \{ \text{AssignmentExpression} \},\\
      \mcode{LeftHandSideExpression} &\mapsto& \{ \text{LeftHandSideExpression} \},\\
      \mcode{AssignmentExpression} &\mapsto& \{ \text{AssignmentExpression} \}
    \end{array}
  \right.
\]
For built-in algorithms, we assign pre-defined variables $\this$, $\args$, and
$\NewTarget$ with corresponding types and parameters with $\tjs$ types.  For
example, the following abstract environment is for the built-in
algorithm $\jscode{Math.round}$ at the upper-right in Figure~\ref{fig:example}:
\[
  \left\{
    \begin{array}{r@{~}c@{~}l}
      \this &\mapsto& \{ \tjs \},
      \args \mapsto \{ \clist{\tjs} \},
      \x \mapsto \tjs\\
      \NewTarget &\mapsto& \{ \text{Object}, \undefval \},\\
    \end{array}
  \right.
\]

\subsection{Condition-based Refinement}\label{sec:refine}

We present a \textit{condition-based refinement} of type analysis for the
modified $\ires$ to increase analysis precision.  It prunes out infeasible parts
in abstract environments based on conditions of branches of assertions, and we
formally define the $\refine$ function as a composition of two functions
$\refineFlow$ and $\refineVar$:
\[
  \refine(\expr, \bool) =
  \refineVar(\expr, \bool) \circ \refineFlow(\expr, \bool)
\]
The $\refineFlow$ prevents infeasible \textit{control flows} by producing an
empty abstract environment as follows:
\[
  \refineFlow(\expr, \bool)(\aenv) = \left\{
    \begin{array}{ll}
      \aenv & \text{if} \; \{ \bool \} \order \aseme{\expr}(\aenv)\\
      \emp & \text{otherwise}\\
    \end{array}
  \right.
\]
Beyond control flows, $\refineVar$ refines the shape of a given abstract
environment based on the conditions for variables:
\[
  \begin{array}{r@{~}c@{~}l}
    \refineVar(\code{!} \expr, \bool)(\aenv) &=&
    \refineVar(\expr, \neg \bool)(\aenv)\\

    \refineVar(\expr_0 \bop \expr_1, \bool)(\aenv) &=&
    \left\{
      \begin{array}{ll}
        \aenv_0 \join \aenv_1 & \text{if} \; \bool \wedge \bop = \code{||}\\
        \aenv_0 \meet \aenv_1 & \text{if} \; \neg\bool \wedge \bop = \code{||}\\
        \aenv_0 \meet \aenv_1 & \text{if} \; \bool \wedge \bop = \code{\&\&}\\
        \aenv_0 \join \aenv_1 & \text{if} \; \neg\bool \wedge \bop = \code{\&\&}\\
        \aenv & \text{otherwise}\\
      \end{array}
    \right.\\
    \multicolumn{3}{r}{
      \text{where} \;
      \aenv_0 = \refineVar(\expr_0, \bool)(\aenv) \wedge
      \aenv_1 = \refineVar(\expr_1, \bool)(\aenv)
    }\\

    \refineVar(\x.\Type \code{=} \nconst{normal}, \true)(\aenv) &=&
    \aenv[ \x \mapsto \aseme{\x}(\aenv) \cap \tnormal(\tyset) ]\\
    \refineVar(\x.\Type \code{=} \nconst{normal}, \false)(\aenv) &=&
    \aenv[ \x \mapsto \aseme{\x}(\aenv) \cap \{ \tabrupt \} ]\\

    \refineVar(\x \code{=} \expr, \true)(\aenv) &=&
    \aenv[ \x \mapsto \aseme{\x}(\aenv) \meet \aseme{\expr}(\aenv) ]\\
    \refineVar(\x \code{=} \expr, \false)(\aenv) &=&
    \aenv[ \x \mapsto \aseme{\x}(\aenv) \setminus
    \lfloor\aseme{\expr}(\aenv)\rfloor ]\\

    \refineVar(\x : \ty, \true)(\aenv) &=&
    \aenv[ \x \mapsto \aseme{\x}(\aenv) \meet \{ \ty \} ]\\
    \refineVar(\x : \ty, \false)(\aenv) &=&
    \aenv[ \x \mapsto \aseme{\x}(\aenv) \setminus \{ \ty' \mid \ty' \subtype \ty
    \}]\\
  \end{array}
\]
and it returns $\aenvset$ for other cases.  The notation $\getSingle{\aty}$ denotes
to checks whether it is a singleton; $\getSingle{\aty} = \aty$ if $\aty = \{ \ty
\}$ and $\ty$ is a singleton type such as a constant $\const$, an empty list
$\clist{}$, an absent value $\tabsent$, a primitive value, etc., and
$\getSingle{\aty} = \emp$ otherwise.

\todo: $\refinePred$?

### 2021-04-13 paper/ase21/checker.tex
[ASE21] Overview
[ASE21] Till 4.2

\section{Static Checker}\label{sec:checker}

\subsection{Reference Checker}
\begin{itemize}
  \item unknown variables
  \item already defined variables
  \item already defined variables
\end{itemize}

\subsection{Parameter Arity Checker}
\begin{itemize}
  \item remaining normal parameters
    (parameter types: normal / optional / variadic)
  \item remaining arguments
\end{itemize}

\subsection{Type Checker}
\begin{itemize}
  \item unchecked abrupt completions
  \item typed operators
\end{itemize}

\subsection{Assertion Checker}
\begin{itemize}
  \item `typeof` assertions
  \item `instance-of` assertions
  \item reachability assertions
\end{itemize}

>>>>

\section{Bug Detector}\label{sec:checker}

\begin{table}
  \centering
  \caption{Type-related specification bugs fixed by pull requests for the recent
  three years from 2018 to 2021.}
  \label{table:pr-bugs}
  \vspace*{-1.5em}
  \[
    \begin{array}{c|l|r}
      \multicolumn{1}{c}{\textbf{Category}} &
      \multicolumn{1}{|c}{\textbf{Bug Kind}} &
      \multicolumn{1}{|c}{\textbf{\# Bugs}}\\
      \hline

      \multirow{2}{*}{\text{Reference}}
      & \text{Unknown Variables} & \inred{XXX}\\\cline{2-3}
      & \text{Already Defined Variables} & \inred{XXX}\\\hline

      \multirow{1}{*}{\text{Parameter}}
      & \text{Arity Mismatches} & \inred{XXX}\\\hline

      \multirow{1}{*}{\text{Assertion}}
      & \text{Assertion Failures} & \inred{XXX}\\\hline

      \multirow{2}{*}{\text{Type}}
      & \text{Non-Numeric Operands} & \inred{XXX}\\\cline{2-3}
      & \text{Unchecked Abrupt Completions} & \inred{XXX}\\\hline

      \multicolumn{2}{c|}{\textbf{Total}} & \inred{XXX}\\

    \end{array}
  \]
  \vspace*{-1.5em}
\end{table}

We develop a \textit{bug detector} to statically detect type-related
specification bugs in ECMAScript using an augmented abstract transfer
$\detector$ with additional checkers.  The abstract semantics $\asem{\prog}$ of
a program $\prog = (\getfunc, \getinst, \getnext)$ is the least fixpoint of the
original abstract transfer $\atransfer$.  Thus, $\detector(\asem{\prog})$ also
produces $\asem{\prog}$ and detects type-related specification bugs.  Before
implementing checkers, we manually investigate pull requests for the recent
\inred{three years from 2018 to 2021}.  As described in
Table~\ref{table:pr-bugs}, pull requests include \inred{XXX} type-related
specification bugs classified in six different kinds and four categories.  We
implement four checkers for each category bugs: \textit{reference checker},
\textit{arity checker}, \textit{assertion checker}, and \textit{operand type
checker}.  Now, we explain how each checker detects corresponding bugs in the
augmented abstract transfer $\detector$.

\subsection{Reference Checker}

In ECMAScript abstract algorithms, variables are dynamically introduced in any
contexts.  The reference bugs are caused when trying to access not yet defined
variables or to redefine already defined variables.  Such reference bugs are the
most prevalent type-related specification bugs according to the our manual
investigation of previous pull requests.  Thus, we implement the reference
checker by adding additional checks to abstract semantics of variable lookups
$\aseme{\x}$ and variable declarations $\asemi{\kwlet \; \x = \expr}$ to detect
unknown variables and already defined variables, respectively:
\[
  \begin{array}{r@{~}c@{~}l}
    \aseme{\x}(\aenv) &=& \left\{
      \begin{array}{ll}
        \text{unknown variable} \; $\x$
        & \text{if} \; \asemr{\x}(\aenv) = \{ \tabsent \}\\
        \cdots
        & \text{otherwise}\\
      \end{array}
    \right.\\

    \asemi{\kwlet \; \x = \expr}(\lab, \tys)(\aelem) &=& \left\{
      \begin{array}{ll}
        \text{already defined variable} \; $\x$
        & \text{if} \; \aty = \{ \true \}\\
        \cdots
        & \text{otherwise}\\
      \end{array}
    \right.\\ &&
    \text{where} \; \aty = \aseme{\x \kwexists}(\aelem(\lab, \tys))\\
  \end{array}
\]
If the abstract semantics of a variable lookup for $\x$ is a singleton $\{
\tabsent \}$, it means the variable $\x$ is always an unknown variable.  In this
case, the reference checker reports a unknown variable bug for $\x$.  For
example, revisit the syntax-directed algorithm at the upper-left in
Figure~\ref{fig:example}.  Since the \textbf{GetReferencedName} algorithm is
removed, the variable \code{GetReferencedName} does not exist in abstract
environments and its lookup always return $\{ \tabsent \}$ thus the reference
checker can detect this unknown variable bug.  For already defined variables,
the reference checker examines whether the variable of each variable declaration
is always defined using the abstract semantics of existence checks $\aseme{\x
\kwexists}$.

\subsection{Arity Checker}
Arity $\size{\func}$ of a function $\func = \kwdef \; \f (\p_1, \cdots, \p_n,
\kwsl \cdots, p_m \kwsr). \; \lab$ is defined as an interval $[n, m]$ where $n$
and $m-n$ denote numbers of normal and optional parameters, respectively.  The
arity checker detects mismatches of such arities with the number of arguments by
adding an additional checks in the function call instructions $\asemi{\x = \kwrl
\expr_0 \; \expr_1 \cdots \expr_k \kwrr}$:
\[
  \begin{array}{l}
    \asemi{\x = \kwrl \expr_0 \; \expr_1 \cdots \expr_k \kwrr}(\lab,
    \tys)(\aelem) =\\
    \qquad \left\{
      \begin{array}{ll}
        \text{missing parameters} \; \p_{k+1}, \cdots, \p_{n_\func} &
        \text{if} \; \exists \func \in \aty_0. \; \text{s.t.} \; k < n_\func\\

        \text{reamining arguments} \; \expr_{m_\func+1}, \cdots, \expr_k &
        \text{if} \; \exists \func \in \aty_0. \; \text{s.t.} \; k > m_\func\\

        \cdots &
        \text{otherwise}\\
      \end{array}
    \right.\\
  \end{array}
\]
where $\func = \kwdef \; \f (\p_1, \cdots, \p_{n_\func}, \kwsl \cdots,
p_{m_\func} \kwsr). \; \lab \wedge \aty_0 = \aseme{\expr_0}(\aelem(\lab,
\tys))$.  For each function $\func$ in the abstract semantics of the function
expression $\expr_0$, the arity checker compares the number of arguments with
its arity $\size{\func}$ and reports a corresponding arity bug.  For example,
consider the lower-left syntax directed algorithm in Figure~\ref{fig:example}.
The algorithm invocation in line 2 is compiled to a function call instruction:
\[
  \x = \kwrl \code{formal}.\code{IteratorBindingInitialization} \; \code{formal}
  \kwrr
\]
with a temporal variable $\x$ thus the number of arguments is 1.  However, arity
of the function stored in $\code{formal}.\code{IteratorBindingInitialization}$
is $[3, 3]$ thus the arity checker reports missing parameter bugs for two
additional parameters $\code{iteratorRecord}$ and $\code{environment}$.

\subsection{Assertion Checker}
\todo

\subsection{Operand Type Checker}
\todo
